master thesis

### 2018-03-20 Paper - Latex/Paper - Skizze - v4.tex
Paper - Methods - rewrite ancestral state reconstruction

  \anmerkung{Mein Vorschlag einer Gliederung (jeweils ca. ein Absatz) (Bernhard)}
  \begin{enumerate}
    \item Motivation:
      \begin{itemize}
        \item Was ist das große Ziel? \\
          Das Ziel dieser Arbeit ist die Anwendung von maximum parsiomony algorithmen auf nicht 
          binäre Bäume und auf sehr große Datensätze. Insbesondere auf das Beispiel 'Entstehung des 
          Parasitismus' im ganzen Eukaryotischen Tree of Life.
        \item Was soll erreicht werden? \\
          Wir wollen vorhandene Algorithm (Sankoff/castor \cite{Louca2017}) auf diese Aufgabenstellung hin testen 
          und ihre Vorhersagekraft abschätzen. Außerdem wollen wir den Fitch algorithmus für binäre 
          Bäume auf unser Problem erweitern und mit dem Sankoff Algorithmus vergleichen.
        \item Warum ist das relevant? Was könnte man dann tun?\\
          predict states of species... \\
          \todo{!!!}
      \end{itemize}
    \item Hintergrund:
      \begin{itemize}
        \item Was gab es in dieser Richtung bereits als ganze Ansätze oder wenn nicht, warum nicht?
          Woran ist es bisher gescheitert? \\
          Bisher wurden vorallem Algorithmen für das binäre Problem entwickelt, da man wesentlich 
            kleinere Teilbäume betrachtet hat, von welchen man auch alle Aufspaltungen kennt. Durch 
            die Entwicklung von OTL, eines gesamten Baum des Lebens, ergibt sich das Problem, dass 
            dieser bei weitem nicht binär ist. \\ \\
  Researchers of the phylogenies have been dealt with the ancenstral state reconstruction in the 
    60s. The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
    Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
    Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
    With more and more data, there is now the possibility to use more information to calculate the 
    probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
    also use branch lengths. The likelihood based algorithms came more in interest. \\
    Our focus came with another 'data extension'. We wanted to work with the biggest phylogenetic tree 
    that exists at this moment, which goes over all observed species. For most \todo{most?} species 
    there is no phylogeny, but only a taxonomic classification.

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{simulating states and transitions between them}
      The next step is to simulate the states of the nodes using the transitions. Again, we simulate 
        fully known states and then 'forget' everything but a few in the leaf nodes so that you can 
        later compare the reconstruction with the origin. \\
      The root node (ancestor of all subsequent species) is (of course) free-living. That means it will 
        start in the free-living beta distribution. Now traverse from the root to the leaf nodes, always 
        pulling out of the current distribution until you get above the threshold and the new node 
        changes state. \\
      To ensure that the parameter of the binomial distribution is restricted to the [0,1] interval, we 
        model it with a beta distribution as in Figure \ref{fig:Beta distribution}. \\

@@

      In the real tree, there is usually only information about species living today $\rightarrow$ leaf 
        nodes. And beyond only a small percentage of these. All information about the states of the 
        internal node and one leaf node is 'forgotten' and stored in another column to the node. \\
      Different percentages of forgetting the information were simulated, as you can read in the 
      \todo{section ... from the results}.

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \section{Ancestral state reconstruction methods}
      For about 50 years, people have been working on ancestral state reconstruction. The first paper 
        mentioned above is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state 
        data \cite{Camin1965}. Different methods have been developed and the question is which method is 
        the most suitable for our problem. \\
      For this purpose, various studied methods and their advantages and disadvantages were compared. \\

@@

      One of the major disadvantages of parsimony methods is that, unlike likelihood approaches, they 
        can not take divergece times (branch length) into account. Since we have no development times in 
        our case, you can ignore this. \\
      Another problem pointed out by Royer-Carenzi is that parsimony approaches are either based on 
        predefined parameters (generalized parsimony) or on strong and often controversial assumptions, 
        like irreversibility for dollo parsimony. Again, this problem is irrelevant for us, because you 
        can only work with generalized models in the analysis of the entire Eukaryota tree. \\

      Following the principle of the simpler model first, the decision has \todo{fallen?} on parsimonious 
        methods since these are sufficient for the case present here. \\
      Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
        previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
        \cite{Farris1970}): Fitch parsiomony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
      \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
        Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
      \todo{Wagner-parsimony \cite{Swofford1987}} \\

@@

          \item Since there are several siblings, do you first of all make the cut or union, or directly 
            in the whole with the father node?

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Implementation}
    You can find the full code on GitHub: 
      \hyperlink{github.com/Irallia/IZW-HU-Parasites}{github.com/Irallia/IZW-HU-Parasites}. \\
    Most of the code was written in Python. The analyzes and the use of the Castor package in R. There 
      are some shell scripts to execute whole workflows.

@@

      For a first overview we collected for every node its number of children (degree $-1$), and plotted
        this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
      The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
        every node (a node with two children is binary). That means it discribes the number of nodes which we have lost from the real (binary) 
      phylogenetic tree. \\
      As you can see, we are very far from a binary tree. \\

@@

      %---------------------------------------------------------------------------------------------------
      \subsubsection{Taxa}
        The investigation of the taxonomy revealed that our tree has three kingdoms: Chloroplastida, 
          Metazoa, Fungi, 53 phyla, 195 classes and 924 orders. \\
        Since the analysis of the tree is not part of this work, it should be mentioned here that, 
          according to recent findings, this is not complete and we lack some taxa in every rank. For 
          example, Cavalier-Smith says that one distinguishes between seven and nine kingdoms 
          \cite{CavalierSmith1981}. \\
        In section \pageref{subsec:listPhyla} of the appendix you can find a list of all phyla.

@@

      As you can see both algortithms are always over 50\% and therefore better than guessing. Moreover, 
        they are usually close to each other, with Sankoff always makes better predictions except for 
        equally distributed states as Fitch. \\
      You can see that the distribution of states has a strong influence on the prediction. The more 
        evenly distributed, the harder it is to predict. \\
      If more than 60\% of internal nodes are missing, Fitch breaks significantly in his prediction 
        compared to Sankoff.

@@

      Weinstein and Kuris have been searching for origins of parasitism in Animalia \cite{Weinstein2016}. 
        They identified 223 parasitic origins: 223 in Metazoa $\supset$ 143 in Arthropoda $\supset$ 87 
        in Insecta. \\
      This has led us to count the origins and losses of parasitism in our investigation as well. \\
      We count only one origin / loss in a parent node with different children's nodes. \\
      Here we have encountered a problem: The Castor algorithm gives us probabilities for states. That 
        means there are also nodes with state like 0.3 or 0.5. So how do you count? Our solution was, 
        to round these values. We have to say that we round 0.5 to 0.

@@

    %---------------------------------------------------------------------------------------------------
    \subsubsection{Distribution of Taxa}
    - In the tree we can distinguish 28 different Taxa with the OTL taxonomic tree. \\
    - The most of them are hardly represented. The major taxonomic groups are: ... \\
    - Here you can see some characteristics of the Multifurcation of the tree. \\
    \begin{figure}[h!]
      \centering
      \includegraphics[width=0.9\textwidth]{Figures/TaxaTable2.JPG}
      \caption{Distribution of Nodes in Rank-Cathegories}
      \label{fig:taxaTable2}
    \end{figure}
    In a phylogeny, the taxonomic division of the tree is far too coarse, meaning that there should 
      be more subtaxa or 'unranked' nodes. But the closer we get to the root, the more the pure
      taxonomic tree is reflected. If the tree were binary, the taxa would have to double. But the 
      multipliers for some are much bigger and for others much smaller, which you can see in in figure 
      \ref{fig:taxaTable2}. \\
      ... (see Table \ref{table:...})

>>>>

  \anmerkung{Mein Vorschlag einer Gliederung (jeweils ca. ein Absatz) (Bernhard)}
  \begin{enumerate}
    \item Motivation:
      \begin{itemize}
        \item Was ist das große Ziel? \\
          Das Ziel dieser Arbeit ist die Anwendung von maximum parsiomony algorithmen auf nicht 
          binäre Bäume und auf sehr große Datensätze. Insbesondere auf das Beispiel 'Entstehung des 
          Parasitismus' im ganzen Eukaryotischen Tree of Life.
        \item Was soll erreicht werden? \\
          Wir wollen vorhandene Algorithm (Sankoff/castor \cite{Louca2017}) auf diese Aufgabenstellung hin testen 
          und ihre Vorhersagekraft abschätzen. Außerdem wollen wir den Fitch algorithmus für binäre 
          Bäume auf unser Problem erweitern und mit dem Sankoff Algorithmus vergleichen.
        \item Warum ist das relevant? Was könnte man dann tun?\\
          predict states of species... \\
          \todo{!!!}
      \end{itemize}
    \item Hintergrund:
      \begin{itemize}
        \item Was gab es in dieser Richtung bereits als ganze Ansätze oder wenn nicht, warum nicht?
          Woran ist es bisher gescheitert? \\
          For about 50 years, people have been working on ancestral state reconstruction. The first paper 
            mentioned above is by Camin and Sokal, who in 1965 were working on algorithms for discrete-state 
            data \cite{Camin1965}. Different methods have been developed and the question is which method is 
            the most suitable for our problem. \\
          For this purpose, various studied methods and their advantages and disadvantages were compared. \\

          Bisher wurden vorallem Algorithmen für das binäre Problem entwickelt, da man wesentlich 
            kleinere Teilbäume betrachtet hat, von welchen man auch alle Aufspaltungen kennt. Durch 
            die Entwicklung von OTL, eines gesamten Baum des Lebens, ergibt sich das Problem, dass 
            dieser bei weitem nicht binär ist. \\ \\
  Researchers of the phylogenies have been dealt with the ancenstral state reconstruction in the 
    60s. The first methods were only brute force \todo{Quelle, siehe Fitch: Camin and Sokal 1965}. 
    Next came a set of parsimony algorithms such as: Fitch-parsimony \cite{Fitch1971}, 
    Wagner-parsimony \cite{Swofford1987} ... \todo{weitere?}. \\
    With more and more data, there is now the possibility to use more information to calculate the 
    probabilities of the ancestral states. In addition to the states of the leafs, algorithms could 
    also use branch lengths. The likelihood based algorithms came more in interest. \\
    Our focus came with another 'data extension'. We wanted to work with the biggest phylogenetic tree 
    that exists at this moment, which goes over all observed species. For most \todo{most?} species 
    there is no phylogeny, but only a taxonomic classification.

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \subsection{simulating states and transitions between them}
      The next step is to simulate the states of the nodes using the transitions. Again, fully known 
        states are simulated and then everything is 'forgotten' but a few in the leaf nodes so that 
        \colorbox{red}{you} can later compare the reconstruction with the origin. \\
      The root node is defined as ancestor of all subsequent species and in this case, determined to be
        free-living. Therefore, a beta distribution for free-living is used at the beginning. Now 
        traverse from the root to the leaf nodes, always pulling out of the current distribution until 
        \colorbox{red}{you} get above the threshold and the new node changes state. \anmerkungstext{Für 
        den Tree/Graph Traversal würde sich eine erklärende Abbildung anbieten. Nur Text macht es sehr 
        schwer verständlich an der Stelle. (Thilo)} \\
      To ensure that the parameter of the binomial distribution is restricted to the [0,1] interval, it 
        is modeled with a beta distribution as in Figure \ref{fig:Beta distribution}. \\

@@

      In the real tree, there is usually only information about species living today $\rightarrow$ leaf 
        nodes. And beyond only a small percentage of these. All information about the states of the 
        internal node and one leaf node is 'forgotten' and stored in another column to the node. \\
      Different percentages of forgetting the information are simulated, as \colorbox{red}{you} can read 
      in the \todo{section ... from the results}.

@@

    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    %---------------------------------------------------------------------------------------------------
    \section{Ancestral state reconstruction methods}
      As presented in the introduction, there are some methods for ancestral state reconstruction. For 
        this purpose, various studied methods and their advantages and disadvantages are compared below. \\

@@

      One of the major disadvantages of parsimony methods is that, unlike likelihood approaches, they 
        can not take divergece times (branch length) into account. Since there are no development times 
        of species in our case, \colorbox{red}{you} can ignore this. \\
      Another problem pointed out by Royer-Carenzi is that parsimony approaches are either based on 
        predefined parameters (generalized parsimony) or on strong and often controversial assumptions, 
        like irreversibility of transitions for dollo parsimony. Again, this problem is irrelevant to 
        the problem at hand, because \colorbox{red}{you} can only work with generalized models in the 
        analysis of the entire Eukaryota tree. \\

      Parsimony-based methods are used in this work, since they are fully sufficient for the presented 
        use case here. Following the principle of the simpler model first. \\
      Felsenstein \cite{Felsenstein2003} discusses in his book two algorithms that generalize all 
        previous methods (from Camin and Sokal \cite{Camin1965}, \todo{Kluge and Farris} and Farris 
        \cite{Farris1970}): Fitch parsiomony \cite{Fitch1971} and Sankoff parsimony \cite{Sankoff1975}. \\
      \anmerkungstext{Unter Farris war auch noch der Begriff Wagner trees in gebrauch, als 
        Verallgemeinerung der parsimonious trees von Camin und Sokal. (Lydia)} \\
      \todo{Wagner-parsimony \cite{Swofford1987}} \\

@@

          \item Since there are several siblings, do \colorbox{red}{you} first of all make the cut or union, 
            or directly in the whole with the father node?

@@

  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  %---------------------------------------------------------------------------------------------------
  \section{Implementation}
    \colorbox{red}{You} can find the full code on GitHub: 
      \hyperlink{github.com/Irallia/IZW-HU-Parasites}{github.com/Irallia/IZW-HU-Parasites}. \\
    Most of the code was written in Python. The analyzes and the use of the Castor package in R. There 
      are some shell scripts to execute whole workflows.

@@

      For a first overview we collected for every node its number of children (degree $-1$), and plotted
        this in two histograms, see figure \ref{fig:childrenOfNodes}. \\
      The multifurcation affects only the internal nodes. We collected the number of children $-2$ of 
        every node (a node with two children is binary). That means it discribes the number of nodes which we have lost from the real (binary) 
      phylogenetic tree. \\
      As \colorbox{red}{you} can see, we are very far from a binary tree. \\

@@

      %---------------------------------------------------------------------------------------------------
      \subsubsection{Taxa}
        The investigation of the taxonomy revealed that our tree has three kingdoms: Chloroplastida, 
          Metazoa, Fungi, 53 phyla, 195 classes and 924 orders. \\
        Since the analysis of the tree is not part of this work, it should be mentioned here that, 
          according to recent findings, this is not complete and we lack some taxa in every rank. For 
          example, Cavalier-Smith says that one distinguishes between seven and nine kingdoms 
          \cite{CavalierSmith1981}. \\
        In section \pageref{subsec:listPhyla} of the appendix \colorbox{red}{you} can find a list of all phyla.

@@

      As \colorbox{red}{you} can see both algortithms are always over 50\% and therefore better than guessing. Moreover, 
        they are usually close to each other, with Sankoff always makes better predictions except for 
        equally distributed states as Fitch. \\
      \colorbox{red}{You} can see that the distribution of states has a strong influence on the prediction. The more 
        evenly distributed, the harder it is to predict. \\
      If more than 60\% of internal nodes are missing, Fitch breaks significantly in his prediction 
        compared to Sankoff.

@@

      Weinstein and Kuris have been searching for origins of parasitism in Animalia \cite{Weinstein2016}. 
        They identified 223 parasitic origins: 223 in Metazoa $\supset$ 143 in Arthropoda $\supset$ 87 
        in Insecta. \\
      This has led us to count the origins and losses of parasitism in our investigation as well. \\
      We count only one origin / loss in a parent node with different children's nodes. \\
      Here we have encountered a problem: The Castor algorithm gives us probabilities for states. That 
        means there are also nodes with state like 0.3 or 0.5. So how do \colorbox{red}{you} count? Our solution was, 
        to round these values. We have to say that we round 0.5 to 0.

@@

    %---------------------------------------------------------------------------------------------------
    \subsubsection{Distribution of Taxa}
    - In the tree we can distinguish 28 different Taxa with the OTL taxonomic tree. \\
    - The most of them are hardly represented. The major taxonomic groups are: ... \\
    - Here \colorbox{red}{you} can see some characteristics of the Multifurcation of the tree. \\
    \begin{figure}[h!]
      \centering
      \includegraphics[width=0.9\textwidth]{Figures/TaxaTable2.JPG}
      \caption{Distribution of Nodes in Rank-Cathegories}
      \label{fig:taxaTable2}
    \end{figure}
    In a phylogeny, the taxonomic division of the tree is far too coarse, meaning that there should 
      be more subtaxa or 'unranked' nodes. But the closer we get to the root, the more the pure
      taxonomic tree is reflected. If the tree were binary, the taxa would have to double. But the 
      multipliers for some are much bigger and for others much smaller, which \colorbox{red}{you} can see in in figure 
      \ref{fig:taxaTable2}. \\
      ... (see Table \ref{table:...})
